Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion

### 2023-07-14 Manuscript.Rmd
updates

---
title: "Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion"
author: 
  - name : "**Jesper Fischer Ehmsen**"
    affiliation: "1"
  - name : "Daniel Elmstrom Christensen"
    affliation : "2"
affliation:

  - id : "1"
    institution : "Center of Functionally Integrative Neuroscience (CFIN), Department of Clinical Medicine, Aarhus University"
  - id : "2"
    institution : "Interacting Minds Center (IMC), Department XXX, Aarhus University"
date: "`r Sys.Date()`"
output:
  word_document:
  html_document: default
  pdf_document:
    latex_engine: xelatex
bibliography: knitting files/Citations.bib
link-citations: yes
linkcolor: blue
always_allow_html: yes
---

@@

The Thermal Grill Illusion (TGI) is a phenomenon in which the juxtaposition of innocuous warm and cold temperatures elicits a burning sensation, offering a unique window to understand how pain can be perceived in response to harmless stimuli. Much debate has revolved around whether spinal mechanisms are involved in the generation of illusory pain, beyond supraspinal mechanisms. In this study, we investigated the role of the spinal cord in the generation of the TGI, in two independent experiments, involving a total of 80 healthy individuals. We applied heat and cold stimuli on dermatomes, namely areas of skin innervated by a single spinal nerve, mapped onto adjacent or nonadjacent spinal segments. Participants were asked to rate their perceptions of cold, warm, and burning sensations in response to TGI and control stimuli. Our aims were to investigate thermosensory and painful perceptual components of the TGI, as well as spatial features of the TGI that may illuminate processes underlying thermosensory integration in the spinal cord. Our findings revealed that both thermosensory and painful components of TGI perception were modulated similarly, with enhanced warm and burning ratings observed when cold and warm stimuli were confined within the same dermatome. Further, we found no perceptual differences based on the proximal-distal location of the cold stimulus within a single dermatome, but a notable enhancement of the TGI effect when the cold stimulus was associated with a more caudal segmental location along the spinal cord compared to the warm stimulus. These results provide insights into the organisation of the spinal cord in relation to the thermosensory integration and generation of the TGI.

@@

The thermal grill illusion (TGI) is a perceptual phenomenon that challenges conventional understanding of pain perception. It is a sensation of burning heat or pain when harmless cold and warm temperatures are applied to the skin simultaneously [@craig_thermal_1994; @craig_functional_1996]. 
Despite cold and warm temperatures being individually innocuous, their combination produces a contradictory burning sensation, even so the temperatures are insufficient to activate peripheral nociceptors. The generation of this illusion is thus attributed to central nervous system mechanisms [@fardo_beyond_2020]. Recent studies have highlighted the involvement of the spinal cord as an initial site contributing to the TGI [@fardo_organization_2018; @harper_conditioned_2017]. However, the precise mechanisms underpinning integration of cold and warm thermal afferents in the spinal cord, alongside those responsible for the distinctive burning quality to this illusion, are yet to be elucidated.The TGI is often described as encompassing two distinct perceptual components - an illusion of heat and an illusion of pain [@defrin_spatial_2008, @fardo_beyond_2020]. The illusion of heat, also known as synthetic heat [@fruhstorfer_significance_2003; @green_localization_1977; -@green_synthetic_2002; -@green_temperature_2004], refers to non-painful sensations evoked by the thermal grill [@defrin_spatial_2008; @kern_pharmacological_2008; -@kern_effects_2008; @bouhassira_investigation_2005; @adam_relationships_2014]. The illusion of pain, which is the most recognised aspect of the TGI, is the distinctive burning sensation that accompanies the simultaneous presentation of cold and warm stimuli [@craig_thermal_1994; -@craig_functional_1996; @bach_thermal_2011]. The hallmark of both illusory components is a qualitative change in perception when the cold and warm stimuli are applied concurrently, compared to when they are presented individually. Historically, the thermosensory and painful components of the TGI were explained by distinct spinal and supraspinal mechanisms, respectively. The enhanced perception of heat in TGI was explained by a spinal inhibitory mechanism, drawing from observations in an animal model where simultaneous application of cold and warm temperatures reduced the activity of cold-specific spinal neurons compared to when cold was applied alone
[@craig_thermal_1994]. Instead, the illusory pain component was ascribed to a disinhibition mechanism at the level of the thalamus, primarily based on the observations of unremitting pain following a thalamic lesion [@craig_thermal_1994; -@bud_craig_new_1998]. Recent human studies on TGI provided differing interpretations of the spinal or supraspinal origin of the illusion. Two studies posited that the illusory pain of the TGI depends uniquely on supraspinal mechanisms. This interpretation was based on the observed modulation of the illusion in accordance with a spatiotopic rather than somatotopic representation of the body [@marotta_transforming_2015]. Further, the illusion remained unaltered during concomitant tactile stimulation, suggesting ineffectiveness of tactile gating - a spinally-mediated process involving inhibition of nociceptive activity by concurrent somatosensory activity [@ferre_ineffectiveness_2018]. Counter to this perspective, other research endorsed a spinal contribution to the TGI. These studies demonstrated that the illusion varied depending on whether cold and warm stimuli were applied to dermatomes mapped either onto adjacent or non-adjacent spinal segments [@fardo_organization_2018]. Participants perceived the stimulation more veridically, consistently with a reduction in TGI perception, when warm or cold stimuli triggered more widespread activity along the spinal cord, corroborating the hypothesis that the spinal cord is an initial site of thermosensory integration underlying TGI. Further support for spinal mechanisms comes from research demonstrating that both noxious heat and the TGI were comparably reduced by conditioned pain modulation in humans. This suggests a similar influence of descending modulation, irrespective of whether the painful sensation was triggered by potentially harmful (noxious) or harmless (innocuous) stimuli [@harper_conditioned_2017]. These findings collectively challenge a purely supraspinal hypothesis of the painful component of the TGI and indicate the relevance of spinal mechanisms in the manifestation of both illusory heat and pain within the TGI.
In this paper our objective was twofold. Firstly, we directly investigated the hypothesis that thermosensory and burning components of the TGI experience are mediated by spinal mechanisms in humans, by manipulating the location of the stimuli within and across dermatomes. Our past work using a similar manipulation involved measuring the experience of the TGI using a temperature matching task [@fardo_organization_2018], which provides a composite measure of TGI perception, reflecting both thermosensory and burning components. Here, to probe possible distinctions between the two qualitative components of the TGI, we measured subjective indices of TGI perception using three independent visual analog scale (VAS) ratings of perceived cold, warm and burning sensations. Secondly, we investigated spatial order effects associated with the integration of cold and warm sensory information at the dermatome (skin) and segmental (i.,e., spine) levels. At the dermatome level, we used body-related coordinates to define proximal (towards the elbow) and distal (towards the wrist) locations. At the spinal level, we used segment-related coordinates to define more rostral (towards the head) and more caudal (towards the lower back) locations. Given the organisation of the spinal cord along a rostral-caudal axis, where each dermatome is represented across multiple spinal segments through the Lissauer tract, this study aimed to glean indirect insights into the spinal mechanisms underpinning thermosensory integration and the generation of the TGI.

\newpage
# Results and discussion
The Thermal Grill Illusion (TGI) is characterised by two key phenomena: thermosensory enhancement and illusory pain. Thermosensory enhancement refers to an amplified perception of heat or cold when cold and warm stimuli are simultaneously applied, as opposed to when each stimulus is presented individually or paired with a neutral temperature. Notably, the majority of individuals experience an intensification of heat rather than cold [@fardo_beyond_2020]. Illusory pain, on the other hand, denotes the perception of a burning sensation elicited by the pairing of warm and cold stimuli, an experience that is largely absent or significantly diminished when each stimulus is presented alone or combined with a neutral temperature. Thus, indicators of a stronger TGI are heightened warm and burning ratings, coupled with reduced cold ratings. 
To investigate thermosensory and burning components of the TGI, participants received pairs of temperatures on their forearms and were asked to quantify the levels of cold, warmth, and burning they experienced during each stimulation. These stimuli consisted of either cold-warm pairs (TGI stimuli), which potentially evoked an illusion of heat and pain, or control stimuli that involved pairing a cold or a warm stimulus of identical temperature as in the TGI condition with a baseline temperature of 30°C (non-TGI stimuli). All stimulation pairs were presented at a fixed distance on the skin, either within the same dermatome or across dermatomes that mapped onto non-adjacent spinal segments. In doing so, the stimuli elicited cold-related and warm-related neural activity that was either more focal or more widespread along the spinal cord. For each stimulation, participants reported their ratings using three sequential VAS scales ranging from 0, indicating the lack of a sensation, to 100, indicating an extreme sensation. For each instance of stimulation, participants were guided to focus their reporting on the sensations originating from a specific thermode. While the participants were unaware, this location corresponded to either the colder stimulus (Exp. 1) or the warmer stimulus (Exp. 2) of the paired temperatures. VAS ratings were analysed using zero-inflated beta regressions.

## Thermosensory and burning components of TGI perception are spinally mediated 
```{r, warning = F, message = F, echo = F,include=FALSE}
if (!require("pacman")) install.packages("pacman")
  pacman::p_load(boot, broom, broom.mixed, brms, car, caret, corrplot, DHARMa, gamlss,
                 ggeffects, gghalves, ggpol, ggpubr, glmmTMB, groupdata2, lme4, lmerTest, 
                 modelsummary, MuMIn, polycor, RColorBrewer, rcompanion, reshape2, Rmisc,
                 rsample, rstatix, simr, tibble, tidyverse, wesanderson, patchwork, cowplot
                 )
source(here::here("scripts","utility.R"))

@@

In keeping with the unique heat and burning features of TGI, our results exhibited a more robust TGI when stimuli were confined within dermatomes compared to when applied across dermatomes, corresponding to non-adjacent spinal segments (Fig. 2). Participants reported a significantly reduced subjective experience of cold (Exp. 1: $\beta$ = `r stats_model_cold_exp1$beta[2]`, `r make_pvalue(stats_model_cold_exp1$p[2])`; Exp. 2: $\beta$ = `r stats_model_cold_exp2$beta[2]`, `r make_pvalue(stats_model_cold_exp2$p[2])`), alongside a markedly enhanced subjective experience of warmth (Exp. 1: $\beta$ = `r stats_model_warm_exp1$beta[2]`, `r make_pvalue(stats_model_warm_exp1$p[2])`; Exp. 2: $\beta$ = `r stats_model_warm_exp2$beta[2]`, `r make_pvalue(stats_model_warm_exp2$p[2])`) for TGI compared to non-TGI stimuli. In addition, a heightened burning sensation was reported in our second experiment (Exp.1: $\beta$ = `r stats_model_burn_exp1$beta[2]`, `r make_pvalue(stats_model_burn_exp1$p[2])`; Exp. 2 $\beta$ = `r stats_model_burn_exp2$beta[2]`, `r make_pvalue(stats_model_burn_exp2$p[2])`) for TGI compared to non-TGI stimuli. These findings not only corroborated earlier results regarding the segmental organisation of the TGI [@fardo_organization_2018] but also suggested that both the thermosensory enhancement and illusory pain components of the illusion are modulated at the spinal cord level. These findings are consistent with previous results on the modulation of heat and pain ratings of TGI stimuli by conditioned pain modulation  [@harper_conditioned_2017]. All together these results support the role of spinal processes in the generation of distinct perceptual aspects of the TGI.

## Undetected proximodistal bias in TGI
Previous research demonstrated a phenomenon known as distal inhibition, wherein heat pain ratings tend to increase when a participant evaluates a more distal compared to a more proximal stimulus among two warm stimuli presented on the forearm [@quevedo_psychophysicshyperalgesia_2004; @quevedo_illusion_2007]. In our study, we sought to investigate whether this effect of distal inhibition would also influence TGI perception within a single dermatome (Fig. 3).
Firstly the cold intensity was less when the cold probe was located more distally for both control and TGI trials (TGI: $\beta$ = `r stats_within_cold_exp1_TGI$beta[2]`, `r make_pvalue(stats_within_cold_exp1_TGI$p[2])`; control: $\beta$ = `r stats_within_cold_exp1_control$beta[2]`, `r make_pvalue(stats_within_cold_exp1_control$p[2])`) with no significant interaction between them  $\beta$ = `r stats_within_cold_exp1_control$beta[5]`, `r make_pvalue(stats_within_cold_exp1_control$p[5])`. No difference in intensity was found for the warm ratings (TGI: $\beta$ = `r stats_within_warm_exp1_TGI$beta[2]`, `r make_pvalue(stats_within_warm_exp1_TGI$p[2])`; control: $\beta$ = `r stats_within_warm_exp1_control$beta[2]`, `r make_pvalue(stats_within_warm_exp1_control$p[2])`) with no significant interaction between them  $\beta$ = `r stats_within_warm_exp1_control$beta[5]`, `r make_pvalue(stats_within_warm_exp1_control$p[5])`. significant differences was also found for the burning ratings indicating more distally located cold probe decreased ratings. (TGI: $\beta$ = `r stats_within_burn_exp1_TGI$beta[2]`, `r make_pvalue(stats_within_burn_exp1_TGI$p[2])`; control: $\beta$ = `r stats_within_burn_exp1_control$beta[2]`, `r make_pvalue(stats_within_burn_exp1_control$p[2])`) with no significant interaction between them  $\beta$ = `r stats_within_burn_exp1_control$beta[5]`, `r make_pvalue(stats_within_burn_exp1_control$p[5])`.

@@

## Directional effects in inter-segmental sensory integration 
A main objective of this study was assessing spatial order effects along the rostrocaudal axis at the spinal level. We delivered an equal number of trials in which the cold stimulus was applied on a dermatome that mapped more rostrally or caudally compared to the warm or neutral stimuli.

@@

The results indicated a notably enhanced TGI effect when the cold stimulus induced more caudal activity within the spinal cord, as depicted in Figure 3. These findings were consistent across both experiments for thermosensory ratings, albeit with minor deviations that corresponded with the particular stimulation quality being assessed. The observed enhancement of warmth perception in Experiment 2 could be ascribed to the participants’ assessment of the warmer thermode as opposed to the colder one of Experiment 1. For burning ratings, significant results were seen in Experiment 2, but not in Experiment 1. One possible reason for this discrepancy could be the relatively low power in Experiment 1, as the power analysis was specifically focused on thermosensory ratings. Alternatively, assessing the warmer thermode might be a more accurate method for measuring TGI perception.

## Spinal organisation and TGI perception
The complexities of spinal neuroanatomy provide insightful perspectives concerning the two main findings of these experiments: (1) enhanced heat and burning sensations when cold-warm thermosensory integration takes place more focally within the spinal cord, and (2) discernible directional inter-segmental effects when distinct cold and warm stimuli elicited broader activity along the spinal cord. Small primary afferents, responsible for mediating temperature and pain sensations, split into ascending and descending branches that cover one to two segments before they enter the dorsal horn [@kerr_neuroanatomical_1975; @lamotte_distribution_1977]. This pattern forms the Lissauer's tract, a structure hypothesised to regulate sensory transmission to the dorsal horn and influence spinal receptive field size [@wall_brief_1999].

@@

# Conclusion
Illusions in the thermo-nociceptive system can be leveraged to improve our understanding of mechanisms contributing to pain perception. Here, we presented results supporting the notion that the spinal cord plays a crucial role in the integration and processing of thermal information, contributing to the perception of both thermosensory enhancement and the illusory pain within the TGI. Additionally, we reported findings on directional inter-segmental effects in spinal integration underlying TGI. Further research is needed to elucidate the neuroanatomical and functional properties of the spinal cord, as well as the intricate interplay between supraspinal and spinal processes, that give rise to TGI perception.

@@

## Participants
The study entailed two separate experiments, collectively involving 80 healthy volunteers. The sample consisted of `r descript_exp1_gender %>% filter(Gender == "F") %>% .$number` females and `r descript_exp1_gender %>% filter(Gender == "M") %>% .$number` males, mean age = `r descript_exp1_age$mean_age` years old (SD = `r descript_exp1_age$sd_age`, range = `r descript_exp1_age$min_age` - `r descript_exp1_age$max_age`) in Experiment 1, and `r descript_exp2_gender %>% filter(Gender == "F") %>% .$number` females and `r descript_exp2_gender %>% filter(Gender == "M") %>% .$number` males and `r descript_exp2_gender %>% filter(Gender != "M" & Gender != "F") %>% .$number` non-binary (Female at birth), mean age = `r descript_exp2_age$mean_age` years old (SD = `r descript_exp2_age$sd_age`, range = `r descript_exp2_age$min_age` - `r descript_exp2_age$max_age`), in Experiment 2. The research methodology complied with the principles set forth in the Declaration of Helsinki and received ethical approval from the Institutional Review Board (IRB) at the Danish Neuroscience Center, Aarhus University, Denmark. Prior to commencing the study, all participants were fully informed about the procedures and provided their voluntary consent.

@@

All thermal stimuli were delivered using two NTE-3 Thermal Sensitivity Testers (PhysiTemp Instruments LLC) controlled by PhysiTemp NTE-3 software (version 5.4b). The procedure involved measurements of heat and cold pain thresholds, calibration of cold-warm temperature pairs eliciting TGI, and an experimental task where TGI and non-TGI stimuli were applied on dermatomes that mapped onto adjacent or non-adjacent spinal segments.

>>>>

---
title: "Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion"
author: 
  - name : "Alexandra G. Mitchell"
    affiliation : "1"
  - name : "Jesper Fischer Ehmsen"
    affiliation : "1"
  - name : "Daniel Elmstrom Christensen"
    affiliation : "2"
  - name : "Anna Villaume Stuckert"
    affiliation : "3"
  - name : "Patrick Haggard"
    affiliation : "4"  
  - name : "Francesca Fardo"
    affiliation : "1"

affiliation:
  - id : "1"
    institution : "Center of Functionally Integrative Neuroscience (CFIN), Department of Clinical Medicine, Aarhus University"
  - id : "2"
    institution : "Interacting Minds Center (IMC), Department XXX, Aarhus University"
date: "`r Sys.Date()`"
output:
  word_document:
  html_document: default
  pdf_document:
    latex_engine: xelatex
bibliography: knitting files/Citations.bib
link-citations: yes
linkcolor: blue
always_allow_html: yes
editor_options: 
  markdown: 
    wrap: sentence
---

@@

The Thermal Grill Illusion (TGI) is a phenomenon in which the juxtaposition of innocuous warm and cold temperatures elicits a burning sensation, offering a unique window to understand how pain can be perceived in response to harmless stimuli.
Much debate has revolved around whether spinal mechanisms are involved in the generation of illusory pain, beyond supraspinal mechanisms.
In this study, we investigated the role of the spinal cord in the generation of the TGI, in two independent experiments, involving a total of 80 healthy individuals.
We applied heat and cold stimuli on dermatomes, namely areas of skin innervated by a single spinal nerve, mapped onto adjacent or nonadjacent spinal segments.
Participants were asked to rate their perceptions of cold, warm, and burning sensations in response to TGI and control stimuli.
Our aims were to investigate thermosensory and painful perceptual components of the TGI, as well as spatial features of the illusion that may illuminate processes underlying thermosensory integration in the spinal cord.
Our findings revealed that both thermosensory and painful components of TGI perception were modulated similarly, with enhanced warm and burning ratings observed when cold and warm stimuli were confined within the same dermatome.
Further, we found no perceptual differences based on the proximal-distal location of the cold stimulus within a single dermatome, but a notable heat enhancement when the cold stimulus was associated with a more caudal segmental location compared to the warm stimulus.
These results provide insights into the organisation of the spinal cord in relation to the thermosensory integration and generation of the TGI.

@@

The thermal grill illusion (TGI) is a perceptual phenomenon that challenges conventional understanding of pain perception.
It is a sensation of burning heat or pain when harmless cold and warm temperatures are applied to the skin simultaneously [@craig_thermal_1994; @craig_functional_1996].
Despite cold and warm temperatures being individually innocuous, their combination produces a contradictory burning sensation, even so the temperatures are insufficient to activate peripheral nociceptors.
The generation of this illusion is thus attributed to central nervous system mechanisms [@craig_thermal_1994; @fardo_beyond_2020].
Recent studies have highlighted the involvement of the spinal cord as an initial site contributing to the TGI [@fardo_organization_2018; @harper_conditioned_2017].
However, the precise mechanisms underpinning integration of cold and warm thermal afferents in the spinal cord, alongside those responsible for the distinctive burning quality to this illusion, are yet to be elucidated.

The TGI is often described as encompassing two distinct perceptual components - an illusion of heat and an illusion of pain [@defrin_spatial_2008, @fardo_beyond_2020]. The illusion of heat, also known as synthetic heat [@fruhstorfer_significance_2003; @green_localization_1977; -@green_synthetic_2002; -@green_temperature_2004], refers to non-painful sensations evoked by the thermal grill [@defrin_spatial_2008; @kern_pharmacological_2008; -@kern_effects_2008; @bouhassira_investigation_2005; @adam_relationships_2014].
The illusion of pain, which is the most recognised aspect of the TGI, is the distinctive burning sensation that accompanies the simultaneous presentation of cold and warm stimuli [@craig_thermal_1994; -@craig_functional_1996; @bach_thermal_2011].
The hallmark of both illusory components is a qualitative change in perception when the cold and warm stimuli are applied concurrently, compared to when they are presented individually.
Historically, the thermosensory and painful components of the TGI were explained by distinct spinal and supraspinal mechanisms, respectively.
The enhanced perception of heat in TGI was explained by a spinal inhibitory mechanism, drawing from observations in an animal model where simultaneous application of cold and warm temperatures reduced the activity of cold-specific spinal neurons compared to when cold was applied alone [@craig_thermal_1994]. Instead, the illusory pain component was ascribed to a disinhibition mechanism at the level of the thalamus, primarily based on the observations of unremitting pain following a thalamic lesion [@craig_thermal_1994; -@bud_craig_new_1998].

Recent human studies on TGI provided differing interpretations of the spinal or supraspinal origin of the illusion.
Two studies posited that the illusory pain of the TGI depends uniquely on supraspinal mechanisms.
This interpretation was based on the observed modulation of the illusion in accordance with a spatiotopic rather than somatotopic representation of the body [@marotta_transforming_2015].
Further, the illusion remained unaltered during concomitant tactile stimulation, suggesting ineffectiveness of tactile gating - a spinally-mediated process involving inhibition of nociceptive activity by concurrent somatosensory activity [@ferre_ineffectiveness_2018].
Counter to this perspective, other research endorsed a spinal contribution to the TGI.
These studies demonstrated that the illusion varied depending on whether cold and warm stimuli were applied to dermatomes mapped either onto adjacent or non-adjacent spinal segments [@fardo_organization_2018].
Participants perceived the stimulation more veridically, consistently with a reduction in TGI perception, when warm or cold stimuli triggered more widespread activity along the spinal cord, corroborating the hypothesis that the spinal cord is an initial site of thermosensory integration underlying TGI.
Further support for spinal mechanisms comes from research demonstrating that both noxious heat and the TGI were comparably reduced by conditioned pain modulation in humans.
This suggests a similar influence of descending modulation, irrespective of whether the painful sensation was triggered by potentially harmful (noxious) or harmless (innocuous) stimuli [@harper_conditioned_2017].
These findings collectively challenge a purely supraspinal hypothesis of the painful component of the TGI and indicate the relevance of spinal mechanisms in the manifestation of both illusory heat and pain within the TGI.

In this paper our objective was twofold.
Firstly, we directly investigated the hypothesis that thermosensory and burning components of the TGI experience are mediated by spinal mechanisms in humans, by manipulating the location of the stimuli within and across dermatomes.
Cold and warm stimuli were presented at a fixed distance on the skin, but depending on their longitudinal or tangential orientation on the arm, they elicited neural activity in a differing number of spinal segments.
Our assumption was that cold and warm-related neural activity in the spinal cord was more focal, when the stimuli were presented within the same dermatome, while more widespread, when the stimuli mapped on spinal segments that were further apart.
Our past work using a similar dermatome manipulation involved measuring the experience of the TGI using a temperature matching task [@fardo_organization_2018], which provides a composite measure of TGI perception, reflecting both thermosensory and burning components.
Here, to probe possible distinctions between the two qualitative components of the TGI, we measured subjective indices of TGI perception using three independent visual analog scale (VAS) ratings of perceived cold, warm and burning sensations.
Secondly, we investigated spatial order effects associated with the integration of cold and warm sensory information at the dermatome (skin) and segmental (i.,e., spine) levels.
At the dermatome level, we used body-related coordinates to define proximal (towards the elbow) and distal (towards the wrist) locations.
At the spinal level, we used segment-related coordinates to define more rostral (towards the head) and more caudal (towards the lower back) locations.
Given the organisation of the spinal cord along a rostral-caudal axis, where each dermatome is represented across multiple spinal segments through the Lissauer tract, this study aimed to glean indirect insights into the spinal mechanisms underpinning thermosensory integration and the generation of the TGI.

# Results and discussion

The Thermal Grill Illusion (TGI) is characterised by two key phenomena: thermosensory enhancement and illusory pain.
Thermosensory enhancement refers to an amplified perception of heat or cold, when cold and warm stimuli are simultaneously applied, as opposed to when each stimulus is presented individually or paired with a neutral temperature.
Notably, the majority of individuals experience an intensification of heat rather than cold [@fardo_beyond_2020].
Illusory pain, on the other hand, denotes the perception of a burning sensation elicited by the pairing of warm and cold stimuli, an experience that is largely absent or significantly diminished when each stimulus is presented alone or combined with a neutral temperature.
Thus, indicators of a stronger TGI are heightened warm and burning ratings, coupled with reduced cold ratings.
To investigate thermosensory and burning components of the TGI, participants received pairs of temperatures on their forearms and were asked to quantify the levels of cold, warmth, and burning they experienced during each stimulation.
These stimuli consisted of either cold-warm pairs (TGI stimuli), which potentially evoked an illusion of heat and pain, or control stimuli that involved pairing a cold or a warm stimulus of identical temperature as in the TGI condition with a baseline temperature of 30°C (non-TGI stimuli).
All stimulation pairs were presented at a fixed distance on the skin, either within the same dermatome or across dermatomes that mapped onto non-adjacent spinal segments.
For each stimulation, participants reported their ratings using three sequential VAS scales ranging from 0, indicating the lack of a sensation, to 100, indicating an extreme sensation.
For each instance of stimulation, participants were guided to focus their reporting on the sensations originating from a specific thermode.
While the participants were unaware, this location corresponded to either the colder stimulus (Exp. 1) or the warmer stimulus (Exp. 2) of the paired temperatures.
VAS ratings were analysed using zero-inflated beta regressions.

## Thermosensory and burning components of TGI perception are spinally mediated

```{r, warning = F, message = F, echo = F,include=FALSE}
if (!require("pacman")) install.packages("pacman")
  pacman::p_load(boot, broom, broom.mixed, brms, car, caret, corrplot, DHARMa, gamlss,
                 ggeffects, gghalves, ggpol, ggpubr, glmmTMB, groupdata2, lme4, lmerTest, 
                 modelsummary, MuMIn, polycor, RColorBrewer, rcompanion, reshape2, Rmisc,
                 rsample, rstatix, simr, tibble, tidyverse, wesanderson, patchwork, cowplot
                 )
source(here::here("scripts","utility.R"))

@@

In keeping with the unique heat and burning features of TGI, our results exhibited a more robust TGI when stimuli were confined within dermatomes compared to when applied across dermatomes, corresponding to non-adjacent spinal segments (Fig. 2).
When rating the cold thermode (Exp. 1), participants reported a significantly reduced subjective experience of cold ($\beta$ = `r stats_model_cold_exp1$beta[2]`, `r make_pvalue(stats_model_cold_exp1$p[2])`), alongside a markedly enhanced subjective experience of warmth ($\beta$ = `r stats_model_warm_exp1$beta[2]`, `r make_pvalue(stats_model_warm_exp1$p[2])`) for TGI stimuli applied within a dermatome compared to across dermatomes.
Burning sensations were enhanced for TGI compared to non-TGI stimuli, but the modulation of the burning ratings by the dermatome condition did not cross the significance threshold ($\beta$ = `r stats_model_burn_exp1$beta[2]`, `r make_pvalue(stats_model_burn_exp1$p[2])`).
In contrast, when rating the warm thermode (Exp. 2), participants reported markedly enhanced burning sensations for TGI stimuli applied within a dermatome compared to across dermatomes ($\beta$ = `r stats_model_burn_exp2$beta[2]`, `r make_pvalue(stats_model_burn_exp2$p[2])`).
However, we did not observe a modulation of cold ($\beta$ = `r stats_model_cold_exp2$beta[2]`, `r make_pvalue(stats_model_cold_exp2$p[2])`) and warm ($\beta$ = `r stats_model_warm_exp2$beta[2]`, `r make_pvalue(stats_model_warm_exp2$p[2])`) thermosensory ratings as in the first experiment.

The findings suggest that the thermosensory quality of the reference thermode (cold or warm) showed differential sensitivity to thermosensory and painful aspects of the TGI experience, and collectively suggested that both the thermosensory enhancement and illusory pain components of the illusion are modulated at the spinal cord level.
This interpretation is consistent with a previous study using a similar dermatome manipulation, but a distinct method to quantify TGI perception [@fardo_organization_2018], as well as another study showing modulation of heat and pain ratings of TGI stimuli by conditioned pain modulation [@harper_conditioned_2017].
All together these results support the role of spinal processes in the generation of distinct perceptual aspects of the TGI.

## Undetected proximodistal bias in TGI

Previous research demonstrated a phenomenon known as distal inhibition, wherein heat pain ratings tend to increase when a participant evaluates a more distal compared to a more proximal stimulus among two warm stimuli presented on the forearm [@quevedo_psychophysicshyperalgesia_2004; @quevedo_illusion_2007].
In our study, we sought to investigate whether this effect of distal inhibition would also influence TGI perception within a single dermatome (Fig. 3).
Firstly the cold intensity was less when the cold probe was located more distally for both control and TGI trials (TGI: $\beta$ = `r stats_within_cold_exp1_TGI$beta[2]`, `r make_pvalue(stats_within_cold_exp1_TGI$p[2])`; control: $\beta$ = `r stats_within_cold_exp1_control$beta[2]`, `r make_pvalue(stats_within_cold_exp1_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_within_cold_exp1_control$beta[5]`, `r make_pvalue(stats_within_cold_exp1_control$p[5])`.
No difference in intensity was found for the warm ratings (TGI: $\beta$ = `r stats_within_warm_exp1_TGI$beta[2]`, `r make_pvalue(stats_within_warm_exp1_TGI$p[2])`; control: $\beta$ = `r stats_within_warm_exp1_control$beta[2]`, `r make_pvalue(stats_within_warm_exp1_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_within_warm_exp1_control$beta[5]`, `r make_pvalue(stats_within_warm_exp1_control$p[5])`.
significant differences was also found for the burning ratings indicating more distally located cold probe decreased ratings.
(TGI: $\beta$ = `r stats_within_burn_exp1_TGI$beta[2]`, `r make_pvalue(stats_within_burn_exp1_TGI$p[2])`; control: $\beta$ = `r stats_within_burn_exp1_control$beta[2]`, `r make_pvalue(stats_within_burn_exp1_control$p[2])`) with no significant interaction between them $\beta$ = `r stats_within_burn_exp1_control$beta[5]`, `r make_pvalue(stats_within_burn_exp1_control$p[5])`.

@@

## Directional effects in inter-segmental sensory integration

A main objective of this study was assessing spatial order effects along the rostrocaudal axis at the spinal level.
We delivered an equal number of trials in which the cold stimulus was applied on a dermatome that mapped more rostrally or caudally compared to the warm or neutral stimuli.

@@

The results indicated a notably enhanced TGI effect when the cold stimulus induced more caudal activity within the spinal cord, as depicted in Figure 3.
These findings were consistent across both experiments for thermosensory ratings, albeit with minor deviations that corresponded with the particular stimulation quality being assessed.
The observed enhancement of warmth perception in Experiment 2 could be ascribed to the participants' assessment of the warmer thermode as opposed to the colder one of Experiment 1.
For burning ratings, significant results were seen in Experiment 2, but not in Experiment 1.
One possible reason for this discrepancy could be the relatively low power in Experiment 1, as the power analysis was specifically focused on thermosensory ratings.
Alternatively, assessing the warmer thermode might be a more accurate method for measuring TGI perception.

## Spinal organisation and TGI perception

The complexities of spinal neuroanatomy provide insightful perspectives concerning the two main findings of these experiments: (1) enhanced heat and burning sensations when cold-warm thermosensory integration takes place more focally within the spinal cord, and (2) discernible directional inter-segmental effects when distinct cold and warm stimuli elicited broader activity along the spinal cord.
Small primary afferents, responsible for mediating temperature and pain sensations, split into ascending and descending branches that cover one to two segments before they enter the dorsal horn [@kerr_neuroanatomical_1975; @lamotte_distribution_1977].
This pattern forms the Lissauer's tract, a structure hypothesised to regulate sensory transmission to the dorsal horn and influence spinal receptive field size [@wall_brief_1999].

@@

# Conclusion

Illusions in the thermo-nociceptive system can be leveraged to improve our understanding of mechanisms contributing to pain perception.
Here, we presented results supporting the notion that the spinal cord plays a crucial role in the integration and processing of thermal information, contributing to the perception of both thermosensory enhancement and the illusory pain within the TGI.
Additionally, we reported findings on directional inter-segmental effects in spinal integration underlying TGI.
Further research is needed to elucidate the neuroanatomical and functional properties of the spinal cord, as well as the intricate interplay between supraspinal and spinal processes, that give rise to TGI perception.

@@

## Participants

The study entailed two separate experiments, collectively involving 80 healthy volunteers.
The sample consisted of `r descript_exp1_gender %>% filter(Gender == "F") %>% .$number` females and `r descript_exp1_gender %>% filter(Gender == "M") %>% .$number` males, mean age = `r descript_exp1_age$mean_age` years old (SD = `r descript_exp1_age$sd_age`, range = `r descript_exp1_age$min_age` - `r descript_exp1_age$max_age`) in Experiment 1, and `r descript_exp2_gender %>% filter(Gender == "F") %>% .$number` females and `r descript_exp2_gender %>% filter(Gender == "M") %>% .$number` males and `r descript_exp2_gender %>% filter(Gender != "M" & Gender != "F") %>% .$number` non-binary (Female at birth), mean age = `r descript_exp2_age$mean_age` years old (SD = `r descript_exp2_age$sd_age`, range = `r descript_exp2_age$min_age` - `r descript_exp2_age$max_age`), in Experiment 2.
The research methodology complied with the principles set forth in the Declaration of Helsinki and received ethical approval from the Institutional Review Board (IRB) at the Danish Neuroscience Center, Aarhus University, Denmark.
Prior to commencing the study, all participants were fully informed about the procedures and provided their voluntary consent.

@@

All thermal stimuli were delivered using two NTE-3 Thermal Sensitivity Testers (PhysiTemp Instruments LLC) controlled by PhysiTemp NTE-3 software (version 5.4b).
The procedure involved measurements of heat and cold pain thresholds, calibration of cold-warm temperature pairs eliciting TGI, and an experimental task where TGI and non-TGI stimuli were applied on dermatomes that mapped onto adjacent or non-adjacent spinal segments.
To measure cold and heat pain thresholds, we gradually adjusted the temperature of one thermode until the participant indicated an experience of pain by pressing a stop button or reached the maximum temperature cut-offs of 5ºC or 50ºC.
We calibrated TGI stimuli by identifying a cold-warm temperature pair based on specific criteria: (1) consistently eliciting a burning sensation of at least 15 on a scale ranging from 0 to 100, (2) consistently avoiding a burning sensation (less than 15) when the cold-neutral (Exp. 1) or warm-neutral (Exp. 2) stimuli were presented, (3) both cold and warm temperatures falling within the innocuous range based on individual cold and heat pain thresholds.
For pain threshold measurements and TGI calibration, we positioned the probes within a single dermatome.
To address our experimental questions, we presented the calibrated TGI stimuli, as well as cold-neutral (Exp. 1) or warm-neutral (Exp. 2) non-TGI stimuli using two thermodes.
In non-TGI stimuli the cold or warm temperatures were set to match the temperature used for TGI stimulation, but paired with a neutral temperature set at 30ºC.
